---
title: "p8105_hw2_sp4170"
author: "Shihui Peng"
date: "2023-10-03"
output: github_document
---
```{r, message=FALSE, echo=FALSE}
library(tidyverse)
```

# Problem 1
First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.
```{r}
pols_month_df =
  read_csv("data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day")) |>
  mutate(month = recode(month, "01" = "January", "02" = "Feburary", "03" = "March", "04" = "April", "05" = "May", "06" = "June", "07" = "July", "08" = "August", "09" = "September", "10" = "October", "11" = "November", "12" = "December")) |> 
  mutate(president = ifelse(prez_gop == 0, "prez_dem", "prez_gop")) |> 
  select(year, month, gov_gop : rep_gop, gov_dem : president)

```
Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.
```{r}
snp_df =
  read_csv("data/snp.csv") |> 
  janitor::clean_names() |> 
  separate(date, into = c("month", "day", "year")) |> 
  mutate(year = as.numeric(year)) |> 
  mutate(year = 2000 + year) |> 
  mutate(month = recode(month, "1" = "January", "2" = "Feburary", "3" = "March", "4" = "April", "5" = "May", "6" = "June", "7" = "July", "8" = "August", "9" = "September", "10" = "October", "11" = "November", "12" = "December")) |> 
  arrange(year, month) |> 
  relocate(year, month) |> 
  select(-day)
```
Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.
```{r}
unemployment_df = 
  read_csv("data/unemployment.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_pct"
  ) |> 
  mutate(month = recode(month, "jan" = "January",  "feb" = "Feburary", "mar" = "March", "apr" = "April", "may" = "May", "jun" = "June", "jul" = "July", "aug" = "August", "sep" = "September", "oct" = "October", "nov" = "November", "dec" = "December"))
```
Join the datasets by merging snp into pols, and merging unemployment into the result.
```{r}
pols_month_df = 
  mutate(pols_month_df, year = as.character(year))

snp_df = 
  mutate(snp_df, year = as.character(year))

unemployment_df = 
  mutate (unemployment_df, year = as.character(year))

pols_month_df = 
  left_join(pols_month_df, snp_df, by = c("year", "month"))

pols_month_df = 
  left_join(pols_month_df, unemployment_df, by = c("year", "month"))
```

