---
title: "p8105_hw2_sp4170"
author: "Shihui Peng"
date: "2023-10-03"
output: github_document
---
```{r, message=FALSE, echo=FALSE}
library(tidyverse)
```

# Problem 1
First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.
```{r}
pols_month_df =
  read_csv("data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day")) |>
  mutate(month = recode(month, "01" = "January", "02" = "Feburary", "03" = "March", "04" = "April", "05" = "May", "06" = "June", "07" = "July", "08" = "August", "09" = "September", "10" = "October", "11" = "November", "12" = "December")) |> 
  mutate(president = ifelse(prez_gop == 0, "prez_dem", "prez_gop")) |> 
  select(year, month, gov_gop : rep_gop, gov_dem : president)

```
Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.
```{r}
snp_df =
  read_csv("data/snp.csv") |> 
  janitor::clean_names() |> 
  separate(date, into = c("month", "day", "year")) |> 
  mutate(year = as.numeric(year)) |> 
  mutate(year = 2000 + year) |> 
  mutate(month = recode(month, "1" = "January", "2" = "Feburary", "3" = "March", "4" = "April", "5" = "May", "6" = "June", "7" = "July", "8" = "August", "9" = "September", "10" = "October", "11" = "November", "12" = "December")) |> 
  arrange(year, month) |> 
  relocate(year, month) |> 
  select(-day)
```
Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.
```{r}
unemployment_df = 
  read_csv("data/unemployment.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment_pct"
  ) |> 
  mutate(month = recode(month, "jan" = "January",  "feb" = "Feburary", "mar" = "March", "apr" = "April", "may" = "May", "jun" = "June", "jul" = "July", "aug" = "August", "sep" = "September", "oct" = "October", "nov" = "November", "dec" = "December"))
```
Join the datasets by merging snp into pols, and merging unemployment into the result.
```{r}
pols_month_df = 
  mutate(pols_month_df, year = as.character(year))

snp_df = 
  mutate(snp_df, year = as.character(year))

unemployment_df = 
  mutate (unemployment_df, year = as.character(year))

pols_month_df_merge = 
  left_join(pols_month_df, snp_df, by = c("year", "month"))

pols_month_df_merge = 
  left_join(pols_month_df, unemployment_df, by = c("year", "month"))
```
Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset.(e.g. give the dimension, range of years, and names of key variables).

Description:

*pols_month_df*:

1. this data set is a `r nrow(pols_month_df)` rows * `r ncol(pols_month_df)` cols data set.
2. the names of variables are `r colnames(pols_month_df)`

*snp_df*:

1. this data set is a `r nrow(snp_df)` rows * `r ncol(snp_df)` cols data set.
2. the names of variables are `r colnames(snp_df)`

*unemployment_df*:

1. this data set is a `r nrow(unemployment_df)` rows * `r ncol(unemployment_df)` cols data set.
2. the names of variables are `r colnames(unemployment_df)`

*Final data set*:

1. the final data set, pols_month_df_merge, is made up of 3 original data sets, which are pols-month.csv, snp.csv, and unemployment.csv. 
2. the range of years contained in the final data set is from 1947 to 2015.
3. the final data set is a `r nrow(pols_month_df_merge)` rows * `r ncol(pols_month_df_merge)` cols data set.
4. the names of variables are `r colnames(pols_month_df_merge)`.

# problem 2

```{r}
library(readxl)
```
Read and clean the Mr. Trash Wheel sheet:

* specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
* use reasonable variable names
* omit rows that do not include dumpster-specific data

The data include a column for the (approximate) number of homes powered. This calculation is described in the Homes powered note, but not applied to every row in the dataset. Update the data to include a new homes_powered variable based on this calculation.
```{r}
mr_trash_wheel_df = 
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N586") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(homes_powered = (500 * weight_tons)/30)
```
Use the similar process to import, clean, and organize the data for Professor Trash Wheel 
```{r}
professor_trash_wheel_df = 
  read_excel("data/202309 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M108") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(homes_powered = (500 * weight_tons)/30)
```

